--- 
title: "A short course on Survival Analysis applied to the Financial Industry"
subtitle: "BBVA Data & Analytics, Madrid"
author: "Marta Sestelo"
date: "27-28 /11/2017, v1.0"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [/Users/Sestelo/Dropbox/bib_tex/bibliografia.bib,/Users/Sestelo/Dropbox/bib_tex/bib_condsurv.bib]
biblio-style: apalike
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "This is a short course on survival analysis applied to the financial field."
---

# Preface {-}



This book is designed to provide a guide for a short course on survival analysis. It is mainly focussed on applying the stastical tecnquines developed in the survival field to the financial industry. The emphasis is placed in understanding the methods, building intuition about when aplying each of them and showing their application through the use of statistical software.





# Programing language and software {-}

The **software** used in the course is the statistical language [`R`](https://cran.r-project.org/) and the IDE (Integrated Development Environment) used is [`RStudio`](https://www.rstudio.com/products/rstudio/download/). A basic prior knowledge of both is assumed. Basic introductions to `R` and `RStudio` are presented in the Appendix \@ref(appendix-r) and \@ref(appendix-rstudio) for those students lacking basic expertise on them.

The required packages for the course are:

```{r, echo = TRUE, cache = TRUE, warning = FALSE, message = FALSE, eval = FALSE}
# Install packages
install.packages(c("survival", "condSURV", "JM", "dplyr", "survminer", "ggplot2"))
devtools::install_github("noramvillanueva/clustcurv")
```

The codes in the notes may assume that the packages have been loaded, so it is better to do it now:
```{r, echo = TRUE, cache = TRUE, warning = FALSE, message = FALSE, eval = TRUE}
# Load packages
library(survival)
library(condSURV)
library(JM)
library(dplyr)
library(survminer)
library(clustcurv)
```

Links: [`survival`](https://cran.r-project.org/web/packages/survival/index.html)[@survival-package],  [`condSURV`](https://cran.r-project.org/web/packages/condSURV/index.html)[@condsurv_package; @meiramachado-sestelo:2016],
[`JM`](https://cran.r-project.org/web/packages/JM/index.html)[@Rizopoulos:2010aa],
[`dplyr`](https://cran.r-project.org/web/packages/dplyr/index.html)[@Wickham:2017aa],
 [`survminer`](https://cran.r-project.org/web/packages/survminer/index.html)[@Kassambara:2017aa],
 [`ggplot2`](https://cran.r-project.org/web/packages/ggplot2/index.html)[@Wickham:2009aa],
 and [`clustcurv`](https://github.com/noramvillanueva/clustcurv).
 



# Main references and credits {-}

Several reference books have been used for preparing these notes. The following list details
the most important ones:

- @kalbfleisch1980statistical
- @book:1205652
- @book:1129209
- @book:1298616
- @kleinbaum2011survival
- @book:1606416

In addition, this material is possible due to the work of persons who contribute greatly to the open source software with incredible pieces of software: @Xie:2015aa, @Xie:2016aa, @R-rmarkdown and @NoRefWorks:1.


The icons used in the notes were designed by [Gregor Cresnar](https://www.flaticon.com/authors/gregor-cresna) from [Flaticon](http://www.flaticon.com/).

All material in these notes is licensed under [CC BY-NC-SA 4.0](https://creativecommons.org/licenses/by-nc-sa/4.0/).


